A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-15 03-semantics.tex
Formatação da sintaxe abstrata na Seção 3
Adicionei macros em header.tex para formatar os comandos da sintaxe
abstrata.  Reescrevi a gramática para usar essas macros.

\newcommand{\NST}{\1\xrightarrow[nst]\1}
\newcommand{\OUT}{\1\xrightarrow[out]\1}
\newcommand{\LL}{\langle}
\newcommand{\RR}{\rangle}
\newcommand{\DS}{\displaystyle}

@@

%\begin{document}

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel highlighting the peculiarities of \CEU, 
in particular the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and system calls (which behave like in conventional 
imperative languages).

@@

\begin{lstlisting}[
  %numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={Reduced syntax of \CEU.},
  label={lst.formal.syntax},
  mathescape=true
]
                                   // primary expressions
  p ::= mem(id)                    (any memory access to `id')
      $|$ awaitExt(id)               (await external event `id')
      $|$ awaitInt(id)               (await internal event `id')
      $|$ emitInt(id)                (emit internal event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ every id p                 (event iteration)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ p @loop p                  (unwinded loop)
      $|$ p @and q                   (unwinded par/and)
      $|$ p @or q                    (unwinded par/or)
      $|$ @canrun(n)                 (can run on stack level `n')
      $|$ @nop                       (terminated expression)
\end{lstlisting}

Listing~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
sufficient to describe all semantic peculiarities of the language.
%
The $mem(id)$ primitive represents all accesses, assignments, and system calls,
that affect a memory location identified by $id$.
%
According to the synchronous hypothesis of \CEU, $mem$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $mem$, $awaitExt$, and $awaitInt$/$emitInt$ expressions do not
share identifiers, i.e., an identifier is either a variable, an external event,
or an internal event.

Regarding mismatches between the concrete and formal languages, the concrete
\code{await} and \code{emit} primitives support communication of values between
them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
\code{v=10}.
To reproduce this functionality in the formal semantics, we can use a shared
variable to hold the value of an \code{emit} and access it after the
corresponding \code{await}.
%
Also, a ``\code{finalize <A> with <B> end ; <C>}'' in the concrete language is
equivalent to ``\code{<A> ; ((fin <B>) or <C>)}'' in the abstract syntax.
In the concrete language, \code{<A>} and \code{<C>} execute in sequence, and
the finalization code \code{<B>} is implicitly suspended waiting for \code{<C>}
termination.
In the formal language, ``\code{fin <B>}'' suspends forever when reached (it is
an awaiting expression that never awakes).
Hence, we need an explicit \code{or} to execute \code{<C>} in parallel, whose
termination aborts ``\code{fin <B>}'', which finally causes \code{<B>} to
execute (by the semantic rules to be discussed).

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from an input event, how the program behaves and 
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program $p$, a stack level $n$, and an
emitted event $e$ to a modified triple as follows:
%
\begin{align*}
\LL p,  n,  e  \RR &\1\xrightarrow\1
\LL p', n', e' \RR
    & \textbf{(rule-general)}
\end{align*}
%
where
%
\begin{align*}
p, p' &\in P
    && (program~as~described~in~Listing~\ref{lst.formal.syntax})
\\
n, n' &\in \mathds{N}
    && (current~stack~level)
\\
e, e' &\in \mathds{E \cup \{\epsilon\}}
    && (emitted~event,~possibly~none)
\end{align*}
%
At the beginning of a reaction to an input event $id$, the triple is
initialized at stack level $0$ ($n=0$) and with the emitted event $id$
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the triple will block with a (possibly) modified program $p'$, at stack level
$0$, with no event emitted $\epsilon$:
%
\begin{align*}
\LL p,0,ext \RR
    & \1\xrightarrow{*}\1
\LL p',0,\epsilon \RR
\end{align*}
%
We also distinguish between two transition rule subtypes:
    \emph{outermost} transitions $\OUT$; and
    \emph{nested} transitions $\NST$.

>>>>

\newcommand{\NST}{\1\xrightarrow[\mathit{nst}]\1}
\newcommand{\OUT}{\1\xrightarrow[\mathit{out}]\1}
\newcommand{\LL}{\langle}
\newcommand{\RR}{\rangle}
\newcommand{\DS}{\displaystyle}

@@

In this section, we introduce a reduced syntax for \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel highlighting the peculiarities of \CEU, 
in particular, the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side-effects and system calls (which behave like in conventional 
imperative languages).

@@

% \begin{lstlisting}[
%   %numbers=left,
%   basicstyle=\ttfamily\footnotesize,
%   float=h,
%   caption={Reduced syntax of \CEU.},
%   label={lst.formal.syntax},
%   mathescape=true
% ]
%                                    // primary expressions
%   p ::= mem(id)                    (any memory access to `id')
%       $|$ awaitExt(id)               (await external event `id')
%       $|$ awaitInt(id)               (await internal event `id')
%       $|$ emitInt(id)                (emit internal event `id')
%       $|$ break                      (loop escape)
%                                    // compound expressions
%       $|$ if mem(id) then p else p   (conditional)
%       $|$ p ; p                      (sequence)
%       $|$ loop p                     (repetition)
%       $|$ every id p                 (event iteration)
%       $|$ p and p                    (par/and)
%       $|$ p or p                     (par/or)
%       $|$ fin p                      (finalization)
%                                    // derived by semantic rules
%       $|$ p @loop p                  (unwinded loop)
%       $|$ p @and q                   (unwinded par/and)
%       $|$ p @or q                    (unwinded par/or)
%       $|$ @canrun(n)                 (can run on stack level `n')
%       $|$ @nop                       (terminated expression)
% \end{lstlisting}

The grammar below defines the syntax of a subset of \CEU that is
sufficient to describe all semantic peculiarities of the language.
\bgroup
\def\lbl#1{\qquad\text{\emph{#1}}}%
\newdimen\X
\X=-1\jot
\begin{alignat*}{2}
  p\Coloneqq
      &\enspace\ceu{\Mem(\Id)}
      &&\lbl{any memory access to~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitExt(\Id)}
      &&\lbl{await external event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitInt(\Id)}
      &&\lbl{await internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\EmitInt(\Id)}
      &&\lbl{emit internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\Break}
      &&\lbl{loop escape}\\[\X]
      %%
  \mid&\enspace\ceu{\Ifelse{b}{p_1}{p_2}}
      &&\lbl{conditional}\\[\X]
      %%
  \mid&\enspace\ceu{p1\,;\,p2}
      &&\lbl{sequence}\\[\X]
      %%
  \mid&\enspace\ceu{\Loop p_1}
      &&\lbl{repetition}\\[\X]
      %%
  \mid&\enspace\ceu{\Every{\Id}\ p_1}
      &&\lbl{event iteration}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\And p_2}
      &&\lbl{par/and}\\[\X]
      %%
  \mid&\enspace\ceu{p_2\Or p_2}
      &&\lbl{par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\Fin p}
      &&\lbl{finalization}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtLoop p_2}
      &&\lbl{unwinded loop}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtAnd\ p_2}
      &&\lbl{unwinded par/and}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtOr\ p_2}
      &&\lbl{unwinded par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\CanRun(n)}
      &&\lbl{can run on stack level~$n$}\\[\X]
      %%
  \mid&\enspace\ceu{\Nop}
      &&\lbl{terminated program}
\end{alignat*}
\egroup

The~$\ceu{\Mem(id)}$ primitive represents all accesses, assignments, and system calls
that affect a memory location identified by~$id$.
%
According to the synchronous hypothesis of \CEU, $\ceu{\Mem}$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and~$\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
share identifiers, i.e., an identifier is either a variable, an external event,
or an internal event.

Regarding mismatches between the concrete and abstract languages, the concrete
\code{await} and \code{emit} primitives support communication of values between
them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
\code{v=10}.
To reproduce this functionality in the formal semantics, we can use a shared
variable to hold the value of an \code{emit} and access it after the
corresponding \code{await}.
%
Also, a ``\code{finalize <A> with <B> end ; <C>}'' in the concrete language is
equivalent to ``\code{<A> ; ((fin <B>) or <C>)}'' in the abstract syntax.
In the concrete language, \code{<A>} and \code{<C>} execute in sequence, and
the finalization code \code{<B>} is implicitly suspended waiting for \code{<C>}
termination.
In the formal language, ``\code{fin <B>}'' suspends forever when reached (it is
an awaiting expression that never awakes).
Hence, we need an explicit \code{or} to execute \code{<C>} in parallel, whose
termination aborts ``\code{fin <B>}'', which finally causes \code{<B>} to
execute (by the semantic rules to be discussed).

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from an input event, how the program behaves and 
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program $p$, a stack level $n$, and an
emitted event $e$ to a modified triple as follows:
%
\begin{align*}
\LL p,  n,  e  \RR &\1\xrightarrow\1
\LL p', n', e' \RR
    & \textbf{(rule-general)}
\end{align*}
%
where
%
\begin{align*}
p, p' &\in P
    && (program~as~described~in~Listing~\ref{lst.formal.syntax})
\\
n, n' &\in \mathbf{N}
    && (current~stack~level)
\\
e, e' &\in \mathbf{E} \cup \{\epsilon\}
    && (emitted~event,~possibly~none)
\end{align*}
%
At the beginning of a reaction to an input event $id$, the triple is
initialized at stack level $0$ ($n=0$) and with the emitted event $id$
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the triple will block with a (possibly) modified program $p'$, at stack level
$0$, with no event emitted $\epsilon$:
%
\begin{align*}
\LL p,0,ext \RR
    & \1\xrightarrow{*}\1
\LL p',0,\epsilon \RR
\end{align*}
%
We also distinguish between two transition rule subtypes:
    \emph{outermost} transitions $\OUT$; and
    \emph{nested} transitions $\NST$.
